\texorpdfstring{Identifying Software Performance Changes Across\\ Variants and Versions}{Identifying Software Performance Changes Across Variants and Versions}

### 2020-08-24 paper.tex
incorporated Sven's CR revisions, still need to revise graphics and a couple of paragrraphs

	%\settopmatter{printacmref=false} 
	%\renewcommand\footnotetextcopyrightpermission[1]{}
	%%
	%% The "title" command has an optional parameter,
	%% allowing the author to define a "short title" to be used in page headers.
	\title{Identifying Software Performance Changes Across Variants and Versions}

@@

	%%
	%% The abstract is a short summary of the work to be presented in the
	%% article.
	\begin{abstract}
		We address the problem of identifying performance changes in the evolution of configurable software systems. Finding optimal configurations and configuration options that influence performance is already difficult, but in light of software evolution, configuration-dependent performance changes may lurk in a potentially large number of different versions of the system.

		In this work, we combine two perspectives -- variability and time --  into a novel perspective. We propose an approach to identify configuration-dependent performance changes retrospectively across software variants and versions. In a nutshell, we iteratively sample pairs of configurations and versions and measure the respective performance that help us update a model of likelihoods for performance changes. Pursuing a search strategy with the goal of measuring selectively and incrementally further pairs, we increase the accuracy of identified change points related to configuration options and interactions.

		We have conducted a number of experiments both on controlled synthetic datasets as well as in real-world scenarios with different software systems. Our evaluation demonstrates that we can pinpoint performance shifts to configuration options and interactions as well as commits introducing change points with high accuracy and at scale. Our experiments on three real-world systems explore the effectiveness and practicality of our approach.

@@

	\section{Introduction}
	%\paragraph{\bfseries What is the problem?}
	Software performance plays a crucial role in users' perception of software quality.
	Overly long execution times, low throughput, or otherwise unexpected performance without added value can render software systems unusable.
	Poor performance is often a symptom of deficiencies in particular software components or the overall software architecture.
	Changes in the observed performance of a software system can be attributed to changes to the software at different levels of granularity (architecture, code, etc.).

	% Das problem sind performance bugs in confi systems
	Typically, modern software systems provide configuration options to enable users and admins to customize behavior to meet different user requirements.
	Configuration options usually correspond to pieces of selectable functionality (features), which contribute to overall performance with different proportions.
	That is, different configurations of a software system exhibit different performance characteristics depending on the configuration decisions.

	Performance changes during software evolution -- intended or not -- can affect all or only a subset of configurations since changes to a software system often relate to a particular configuration option or set of options.
	This is why performance bugs are rarely visible in default configurations, but revealed only in certain configurations \cite{hanEmpiricalStudyPerformance2016}.
	If undetected, such perennial bugs can persist for the lifetime of a software system as software evolves.
	Adding this \emph{technical debt} constantly can accumulate and entail trends of degrading performance quality~\cite{guoGeneticAlgorithmOptimized2011}.

	Performance assessment and estimation is a resource-intense task with many possible pitfalls prevailing.
	Best practices for conducting performance measurements emphasize a dedicated and separated hardware setup to prevent measurement bias by side-processes and, subsequently, obtain reproducible results.	% performance measurement
	What is often overlooked is the fact that most software systems are configurable which introduces another layer of complexity.
	Due to combinatorics, even for a small number of configuration options, the range of possible valid configurations renders exhaustive measurements infeasible.
	%Most problems involving configuration options, such as finding a configuration with optimal performance~\cite{ohFindingNearoptimalConfigurations2017,nairFlash18} or estimating the influence on performance of an individual option~\cite{siegmundPredictingPerformanceAutomated2012,siegmundPerformanceinfluenceModelsHighly2015,sarkarCostEfficientSamplingPerformance,haDeepPerf2019}.

@@

	Our goal is to enable the \emph{retrospective} detection of changes in the performance of configurable software systems and pinpoint them to a specific option or interaction. For example, a patch of a certain feature is likely to affect only configurations where such feature is selected. Many software systems are configurable, but have no performance regression testing routine set in place. To uncover performance deficiencies that emerged from revisions in the past development history, it is infeasible to test each and every commit and configuration. In essence, we face a huge combinatorial problem for performance-change detection: code changes across software versions and behavioral changes across software variants.
	Our problem arises along two dimensions: time (versions) and configuration space (variants).
	\emph{First}, we aim at finding performance changes in software performance from one version to another.
	The detection of such changes is referred to as \emph{change point detection} and is a recurring theme in temporal analysis.
	The main limiting factor is that exhaustive measurements across the configuration space are neither available nor feasible~\cite{white_selecting_2009}.
	Change point detection techniques with both exhaustive measurement~\cite{cityIdentifying2014,daly_industry_2020} as well as limited data availability~\cite{sandoval_alcocer_learning_2016,alcocer_prioritizing_2020,huang_performance_2014,muhlbauer_accurate_2019} have been successfully applied to identify performance changes.
	\emph{Second}, we aim at associating changes in performance (from one version to another) with particular configuration options or interactions among them.
	There is substantial work  regarding a related problem, which is estimating the influence of individual options on performance~\cite{siegmundPredictingPerformanceAutomated2012,siegmundPerformanceinfluenceModelsHighly2015,sarkarCostEfficientSamplingPerformance,haDeepPerf2019}. Instead of estimating the influence of options and interactions on performance, we want to know: Which option or interaction is \emph{responsible} for a particular change point?

	%\begin{itemize}
	%	\item Work so far has only considered performance from a configuration perspective or a evolution perspective.
	%	\item Temporal analysis and general comparison of performance measurements is mostly well-understood. Mühlbauer et al estimate performance of single variants using Gaussian Process Models, yet
	%\end{itemize}

	%\paragraph{\bfseries What are the key components of my approach and results?}
	%Our main idea is as follow: we describe a model to conceive {\color{bostonred}change in individual variants’ performance history across a software’s existing lifetime.}
	%For a representative number of variants, we estimate the probability distribution for each commit to exhibit an shift in performance. Only variants, whose configurations share common features or interactions are affected by a change point.
	%Based on the estimated probability distributions for each variant, we estimate the conditional probability for each commit to be a change point, given a certain set of options is selected.
	%The outcome is a model that indicates (1) where performance changes and (2) which configurations are likely to be affected by change points, both without exhaustive measurements.
	%Subsequently, we present an adaptive extension to our approach that lets us iteratively and efficiently learn accurate prediction models for

	Our main idea is as follows: We address the configuration complexity of this problem by selecting representative sample sets of configurations.
	Next, we uniformly sample a constant number of commits for each configuration and conduct respective performance measurements.
	Based on these measurement data, we assemble a prediction model.
	For each configuration, we estimate the likelihood of each commit being a change point, i.e. that performance of some configurations changes abruptly compared to the previous commit.
	Next, we leverage similarities in the performance histories of configurations that share common options.
	Since we sample the commits for each configuration independently, for each change point, we obtain many estimations using measurements of different commits.
	Overall, this allows us to obtain more accurate estimations of performance-changing commits with tractable effort.
	Based on a mapping from configurations to predicted change point probabilities for each configuration and commit, we derive the options responsible for each particular change point: configuration-dependent change points.
	In summary, we offer the following contributions:
	\begin{itemize}
		\item A novel technique to effectively identify shifts in performance of configurable software systems. We are able to pinpoint causative commits and affected configuration options with high accuracy and tractable effort.
		\item A feasibility demonstration of our approach by implementing an adaptive learning strategy to obtain accurate estimations with acceptable measurement cost.
		\item An evaluation using both synthetic and real-world performance data from three configurable software systems. Synthetic data lets us assess our model and approach conceptually and at scale, whereas we are able to assess practicality with real-world data.
		\item A companion website\footnote{\url{https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions} or archived at \url{https://archive.softwareheritage.org/browse/origin/https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions}} providing supplementary material including a reference implementation of our approach, performance measurement data, and additional visualizations.
	\end{itemize}

	\section{Configuration-Dependent Change Points}
	Performance as a property, such as execution time, emerges from a variety of factors.
	Besides external factors, including hardware setup or execution environment~\cite{ousterhout_always_2018}, the configuration  of a software system can influence performance to a large extent~\cite{siegmundPredictingPerformanceAutomated2012}.
	Most modern software systems exhibit configuration options that correspond to selectable pieces of functionality.
	With configuration options, we turn features on and off creating a variant of the software system.
	Depending on the configuration, different system variants with different behavior and performance can be derived.

	\subsection{Performance-Influence Models}\label{sec:pimodels}
	Consider the following running example of a hypothetical database management system (DBMS) with two selectable features: \textsc{Encryption} and \textsc{compression}.
	Either feature adds execution time to the overall performance, but, if both features are selected, the execution time is smaller than the sum of the individual features' contribution to performance: Less data is encrypted if it is compressed beforehand.
	An interaction in this setting is the combined effect of features \textsc{Encryption} and \textsc{compression}.
	We can assign each feature and interaction an influence it contributes to the overall performance, if selected. In our example, the individual influences of the two features are positive (increasing execution time), whilst the interaction's influence is negative (decreasing execution time).

	The influence of features on performance can be described using \emph{performance-influence models}~\cite{siegmundPerformanceinfluenceModelsHighly2015}.
	A performance-influence model is a linear prediction model of the form $\Pi:C\rightarrow \mathbb{R}$, whereby $C$ denotes a configuration vector (assignment of concrete values to configuration options) and the estimate is a system variant's real-valued performance.
	A formal representation is as follows.

	In Equation~\ref{eq:pimodel}, $F$ denotes the set of all features. Our linear model has $\vert 2^F\vert$ terms, whereby each term $t \subseteq F$ corresponds to a subset of $F$. Each term is described by a coefficient $\beta_f$ and a configuration parameter $c_f$.
	These coefficients encode the performance-influence of features and interactions. $c$ is a vector and denotes the assignment of concrete values to configuration options or interactions. A configuration parameter $c_t$ evaluates to $1$, if all configuration options of the corresponding subset $t \subseteq F$ are selected, and $0$ otherwise. The empty set corresponds to the invariable core functionality. The singleton subsets correspond to individual features, compound subsets to interactions of features.

	\begin{align} \label{eq:pimodel}
	\begin{split}
	\Pi(c) = \beta^T c =  \begin{bmatrix}
	\beta_1\\
	\vdots \\
	\beta_{2^{\vert F\vert}}
	\end{bmatrix}^T \cdot \begin{bmatrix}
	c_1\\
	\vdots \\
	c_{2^{\vert F\vert}}
	\end{bmatrix}
	\end{split}
	\begin{split}
	\beta_i \in \mathbb{R}^{2^{|F|}}\\
	c_i \in \left[0,1\right]^{2^{|F|}}
	\end{split}
	\end{align}

	For our DBMS example, we present a fully determined performance-influence model in Equation~\ref{eq:pimodel_dbms}. The first two terms correspond to the features \textsc{Encrpytion} and \textsc{Compression}, respectively. The third term respresents the interaction of both features. The configuration parameter $c_{\textsc{Compr}\land\textsc{Encrypt}}$ evaluates to $1$, if both $c_{\textsc{Encrypt}}$ and $c_{\textsc{Compr}}$ evaluate to $1$.
	That is, $c_{\textsc{Compr}\land\textsc{Encrypt}} = c_{\textsc{Encrypt}}c_{\textsc{Compr}}$. In the last term, we omit the configuration parameter $c_\emptyset$, as this term represents invariable functionality.

@@

	Prediction models of configuration-dependent performance of this form are not the only possible representation~\cite{guoVariabilityawarePerformancePrediction2013,haDeepPerf2019}, but linear models are easily explainable because they represent the performance-influence of features and interactions as the linear model's coefficients.
	We review extensions and alternatives to linear performance-influence models in Section~\ref{sec:rw}.

	\subsection{Evolution of Performance Influences}
	Performance-influence models describe how features shape performance, but they do not allow for understanding configuration-dependent performance changes over time.
	We expand our DBMS example with a temporal dimension, considering a development history of hundred commits.
	In particular, we assume the following three events. (1) At commit 25, feature \textsc{Compression} has been modified, increasing the execution time; feature \textsc{Encryption} is introduced and can be combined with \textsc{Compression}. (2) At commit 50, the interaction of \textsc{Compression} and \textsc{Encryption} changes, resulting in an increased execution time. (3) At commit 80, the core functionality of the DBMS is refactored, which decreased execution time for all variants.
	\begin{figure}[ht!]
		%		\begin{subfigure}{\linewidth}
		\centering
		% include first image
		\includegraphics[width=\linewidth]{images/dbms_performance.eps}
		\caption{Performance of 4 variants with 3 change points}
		\label{fig:dbms_performance}
	\end{figure}

	The performance histories of valid four variants in Figure~\ref{fig:dbms_performance} exhibit three change points at commits 25, 50, and 80. These change points, however, do not affect all configurations.
	For  instance, execution time decreases for all configurations at commit 80, but increases at commit 50 only for one configuration.
	Given this example of performance histories, the target outcomes of our approach are (1) the locations of the three change points at commits 25, 50, and 80, and (2) the association of such commits to the features/interactions \textsc{Compr}, \textsc{Compr\ $\land$\ Encrpyt}, and the invariable functionality, respectively.

@@

	\subsection{Taming Complexity}
	A naive approach for identifying change points in the influence of different options and interactions on performance over time is to combine existing work on performance-prediction over time and across software variants.

	For our example, we could measure all variants for each commit building a performance-influence model per commit.
	Having 4 configurations and 100 commits, this would result in 400 measurements.
	Since the number of configurations grows exponentially with the number of features, we end up with $2^n$ times $T$ measurements where $n$ is the number of features and $T$ is the number of commits.
	Clearly, this does not scale along both dimensions: Already a few features would render even small commit histories intractable and even limited configurability in the presence of realistic commit histories of a few thousands of commits make this approach infeasible.

@@

	Considering the introductory DBMS example (cf. Figure~\ref{fig:dbms_performance}), we illustrate our sampling strategy in Figure~\ref{fig:sampling plan}. Here, each box depicts a possible measurement, i.e., a pair of a configuration and a commit. The black bars represent change points hidden in the software system‘s performance histories, i.e., measurements immediately before and after a change point are dissimilar. The current state of the sample set comprises all measurements that are filled in grey.

	To include new measurements into the sample set, we take into account four possible objectives. On the one hand, our sample set might already hint to some possible change points and associated configuration options. On the other hand, our sample set might be too sparse and we cannot infer some change points yet. To address this trade-off, we split our objectives into exploitation and exploration of configurations and commits, respectively.

	 For \emph{exploitation}, we sample new measurements to verify guesses based on the current state. That is, in Figure~\ref{fig:sampling plan}, we might include the measurements depicted by black-filled symbols~($\starletfill$/$\pentagofill$). The black-filled stars~($\starletfill$) represent measurements are of interest because we have already identified one change point for the configuration \textsc{Compression}. We exploit this knowledge to test for further configuration options involved in this change point. The black-filled pentagons~($\pentagofill$) represent measurements of interest, because we have identified another early change point for \textsc{Compression}, but within a rather broad range of commits. Thus, we include further measurements from that range to narrow down the possible range for this particular change point.

	 For \emph{exploration}, we might include the measurements by non-filled symbols~($\starlet$/$\pentago$). The non-filled stars~($\starlet$) represent measurements that explore new configurations. In our example, the interaction between \textsc{Compression} and \textsc{Encryption} is the only configuration left to be measured. We include several commits of a particular new configuration to unveil possible performance variation that indicate possible change points. The non-filled pentagons ($\pentago$) represent measurements that are of interest because they  increase the overall measurement coverage. We include measurements from large intervals of not yet measured commits since possible change points can be hidden here.

	These four strategies (exploration and exploitation for configurations and commits) prioritize measurements to be included next into the sample set. Adaptive sampling techniques have been successfully applied to obtain both performance-influence models~\cite{siegmundPredictingPerformanceAutomated2012,sarkarCostEfficientSamplingPerformance} and performance histories~\cite{muhlbauer_accurate_2019} before. However, it is unclear whether fewer measurements are sufficient to assess the performance-influence of configuration options and interactions with respect to version changes. This is what we aim to accomplish with our work.

@@

	\section{An algorithm for change point detection}\label{sec:cpmodel}
	We propose an algorithm to detect substantial shifts in performance of configurations and associate them to individual commits and options or interactions. We lay out our approach as an iterative search across the commit history and configuration space.

@@

	Subsequently, our algorithm estimates a \emph{candidate solution}, which is a set of pairs of a commit and a configuration option. 
	Each of such pairs describes the estimated involvement of a configuration option in the shift of performance influence. 
	Interactions are conceived as multiple tuples with identical commits. That is, if a commit occurs in multiple tuples, each with different configuration options, this indicates that the shift in a performance influence arising from an interaction among two or more options. 
	In the following, we will refer to such pairs as \emph{associations}.

	After obtaining one candidate solution per search iteration, we augment the sample set of measurements with regard to two objectives: exploration and exploitation. 
	\emph{Exploration} aims at including previously unseen commits and configurations to improve coverage of the search space. 
	\emph{Exploitation} aims at including measurements in the sample set that, based on the previous candidate solution, may increase confidence in associations or rule out false positives (i.e., commits falsely identified as a change point or options falsely associated with a commit). 
	For the latter objective, we make an informed decision of what measurements are to be included, whereas the first objective is agnostic of previous candidate solutions.

	As each iteration yields a candidate solution, we keep track of associations in a \emph{solution cache} and repeatedly update the confidence of associations. 
	The rationale is not to lose previously identified change points, but at the same time, allow for removing identified changes points that are likely false positives due to new measurements.
	Some associations, especially in the beginning, might be influenced by sampling bias and can be removed if successive iterations do not repeatedly revisit these associations. 
	The algorithm terminates if the solution cache does not change for a number of iterations or a maximum number of iterations/measurements has been reached.
	Next, we present the steps of our approach in detail.

	\begin{figure}[ht]
		\includegraphics[width=\linewidth]{images/diagram_april.pdf}
		\caption{Overview of our approach:  \textcircled{1} Selecting training data across configuration space and time, \textcircled{2} estimating the change point probability distribution per configuration, \textcircled{3} estimating change points as candidate solutions, \textcircled{4} adaptively augmenting the training set, and \textcircled{5} updating the solution cache and evaluating termination.}
		\label{fig:overview}
	\end{figure}

>>>>

	%\settopmatter{printacmref=false} 
	%\renewcommand\footnotetextcopyrightpermission[1]{}
	%%
	%% The "title" command has an optional parameter,
	%% allowing the author to define a "short title" to be used in page headers.
	\title[Identifying Software Performance Changes Across Variants and Versions]{\texorpdfstring{Identifying Software Performance Changes Across\\ Variants and Versions}{Identifying Software Performance Changes Across Variants and Versions}}

@@

	%%
	%% The abstract is a short summary of the work to be presented in the
	%% article.
	\begin{abstract}
		We address the problem of identifying performance changes in the evolution of configurable software systems. Finding optimal configurations and configuration options that influence performance is already difficult, but in the light of software evolution, configuration-dependent performance changes may lurk in a potentially large number of different versions of the system.

		In this work, we combine two perspectives---variability and time---into a novel perspective. We propose an approach to identify con\-figuration-dependent performance changes retrospectively across the software variants and versions of a software system. In a nutshell, we iteratively sample pairs of configurations and versions and measure the respective performance, which we use to update a model of likelihoods for performance changes. Pursuing a search strategy with the goal of measuring selectively and incrementally further pairs, we increase the accuracy of identified change points related to configuration options and interactions.

		We have conducted a number of experiments both on controlled synthetic datasets as well as in real-world scenarios with different software systems. Our evaluation demonstrates that we can pinpoint performance shifts to individual configuration options and interactions as well as commits introducing change points with high accuracy and at scale. Experiments on three real-world systems explore the effectiveness and practicality of our approach.

@@

	\section{Introduction}
	%\paragraph{\bfseries What is the problem?}
	Software performance plays a crucial role in users' perception of software quality.
	Excessive execution times, low throughput, or otherwise unexpected performance without added value can render software systems unusable.
	Poor performance is often a symptom of deficiencies in particular software components or the overall software architecture.
	Changes in the observed performance of a software system can be attributed to changes to the software at different levels of granularity (architecture, code, etc.).

	% Das problem sind performance bugs in confi systems
	Typically, modern software systems provide configuration options to enable users and admins  customizing behavior to meet different user requirements.
	Configuration options usually correspond to pieces of selectable functionality (features), which contribute to overall performance with different proportions.
	That is, different configurations of a software system exhibit different performance characteristics depending on configuration decisions made by the user.

	Performance changes during software evolution---intended or not---can affect all or only a subset of configurations since changes to a software system often relate to a particular configuration option or set of options.
	This is why performance bugs are rarely visible in default configurations, but revealed only in certain configurations~\cite{hanEmpiricalStudyPerformance2016}.
	If undetected, such perennial bugs can persist for the lifetime of a software system as software evolves.
	Adding this \emph{technical debt} constantly can accumulate and entail trends of degrading performance quality~\cite{guoGeneticAlgorithmOptimized2011}.

	Performance assessment and estimation is a resource-intense task with many possible pitfalls prevailing.
	Best practices for conducting performance measurements emphasize a dedicated and separated hardware setup to prevent measurement bias by side-processes and, subsequently, obtain reproducible results~\cite{molyneauxArtApplicationPerformance2015}.	% performance measurement
	What is often overlooked is the fact that most software systems are configurable, which introduces another layer of complexity.
	Due to combinatorics, even for a small number of configuration options, the range of possible valid configurations renders exhaustive measurements infeasible.
	%Most problems involving configuration options, such as finding a configuration with optimal performance~\cite{ohFindingNearoptimalConfigurations2017,nairFlash18} or estimating the influence on performance of an individual option~\cite{siegmundPredictingPerformanceAutomated2012,siegmundPerformanceinfluenceModelsHighly2015,sarkarCostEfficientSamplingPerformance,haDeepPerf2019}.

@@

	Our goal is to enable the \emph{retrospective} detection of changes in the performance of configurable software systems and pinpoint them to a specific option or interaction. For example, a patch of a certain feature is likely to affect only configurations where such feature is selected. 
	We would like to know in which revision this patch was introduced and which features were affected.
	Many software systems are configurable, but have no performance regression testing routine set in place. 
	To uncover performance deficiencies that emerged from revisions in the past development history, it is infeasible to test each and every commit and configuration. 
	In essence, we face a combinatorial explosion along two dimensions: time (versions) and configuration space (variants).
	\emph{First}, we aim at finding changes in software performance from one version to another.
	The detection of such changes is referred to as \emph{change point detection}~\cite{daly_industry_2020,cityIdentifying2014,huang_performance_2014,muhlbauer_accurate_2019,sandoval_alcocer_learning_2016,alcocer_prioritizing_2020}.
	The main limiting factor is that exhaustive measurements across the configuration space are neither available nor feasible~\cite{white_selecting_2009}.
	Change point detection techniques with both exhaustive measurement~\cite{cityIdentifying2014,daly_industry_2020} as well as limited data availability~\cite{sandoval_alcocer_learning_2016,alcocer_prioritizing_2020,huang_performance_2014,muhlbauer_accurate_2019} have been successfully applied to identify performance changes.
	\emph{Second}, we aim at associating performance changes (from one version to another) with particular configuration options or interactions among them.
	There is substantial work  regarding a related problem, which is estimating the influence of individual options on performance~\cite{siegmundPredictingPerformanceAutomated2012,siegmundPerformanceinfluenceModelsHighly2015,sarkarCostEfficientSamplingPerformance,haDeepPerf2019} for a single version of the software system, ignoring the temporal dimension. Instead of estimating the influence of options and interactions on performance, we want to know: Which option or interaction is \emph{responsible} for a particular change point in the version history of a software system?

	Our main idea is as follows: We address the configuration complexity of this problem by selecting representative sample sets of configurations.
	Then, we uniformly sample a constant number of commits for each configuration and conduct respective performance measurements.
	Based on these measurements, we learn a prediction model.
	For each configuration, we estimate the likelihood of each commit being a change point (i.e., that performance of some configurations changes abruptly compared to the previous commit).
	Next, we leverage similarities in the performance histories of configurations that share common options.
	Since we sample the commits for each configuration independently, we obtain for each change point many estimations using measurements of different commits.
	Overall, this allows us to obtain more accurate estimations of performance-changing commits with tractable effort.
	Based on a mapping from configurations to predicted change point probabilities for each configuration and commit, we derive the options responsible for each particular change point, which we call \emph{configuration-dependent change points}. 	
	In summary, we offer the following contributions:
	\begin{itemize}
		\item A novel technique to effectively identify shifts in performance of configurable software systems. It is able to pinpoint causative commits and affected configuration options with high accuracy and tractable effort.
		\item A feasibility demonstration of our approach by implementing an adaptive learning strategy to obtain accurate estimations with acceptable measurement cost.
		\item An evaluation using both synthetic and real-world performance data from three configurable software systems. Synthetic data let us assess our model and approach conceptually and at scale, whereas we are able to assess practicality with real-world data.
		\item A companion Web site\footnote{\url{https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions/} or an archived version at \url{https://archive.softwareheritage.org/browse/origin/https://github.com/AI-4-SE/Changepoints-Across-Variants-And-Versions/}} providing supplementary material including a reference implementation of our approach, performance measurement data, and additional visualizations.
	\end{itemize}

	\section{Configuration-Dependent Change Points}
	Performance as a property emerges from a variety of factors.
	Besides external factors, including hardware setup or execution environment~\cite{ousterhout_always_2018}, the configuration  of a software system can influence performance to a large extent~\cite{siegmundPredictingPerformanceAutomated2012}.
	Most modern software systems exhibit configuration options that correspond to selectable pieces of functionality.
	With configuration options, we turn features on and off creating a variant of the software system.
	Depending on the configuration, different system variants with different behavior and performance can be derived.

	\subsection{Performance-Influence Models}\label{sec:pimodels}
	Consider the following running example of a database management system (DBMS) with two selectable features: \textsc{Encryption} and \textsc{compression}.
	Either feature adds execution time to the overall performance, but, if both features are selected, the execution time is smaller than the sum of the individual features' contribution to performance: Less data is encrypted if it is compressed beforehand.
	An \textit{interaction} in this setting is the combined effect of features \textsc{Encryption} and \textsc{compression}.
	We can assign each feature and interaction an \textit{influence} that it contributes to the overall performance, if selected. In our example, the individual influences of the two features are positive (increasing execution time), whilst the interaction's influence is negative (decreasing execution time).

	The influence of features on performance can be described using a \emph{performance-influence model}~\cite{siegmundPerformanceinfluenceModelsHighly2015}.
	A performance-influence model is a linear prediction model of the form $\Pi:C\rightarrow \mathbb{R}$, whereby $C$ denotes a configuration vector (assignment of concrete values to configuration options) and the estimate is a system variant's real-valued performance:

	\begin{align} \label{eq:pimodel}
		\begin{split}
		\Pi(c) = \beta^T c =  \begin{bmatrix}
		\beta_1\\
		\vdots \\
		\beta_{2^{\vert F\vert}}
		\end{bmatrix}^T \cdot \begin{bmatrix}
		c_1\\
		\vdots \\
		c_{2^{\vert F\vert}}
		\end{bmatrix}
		\end{split}
		\begin{split}
		\beta_i \in \mathbb{R}^{2^{|F|}}\\
		c_i \in \lbrace 0,1\rbrace^{2^{|F|}}
		\end{split}
	\end{align}

	$F$ denotes the set of all features. Our linear model has $\vert 2^F\vert$ terms, whereby each term $t \subseteq F$ corresponds to a subset of $F$ and is described by a coefficient $\beta_f$ and a configuration parameter $c_f$.
	Coefficients encode the performance-influence of features and interactions. $c$ is a vector and denotes the assignment of concrete values to configuration options or interactions. A configuration parameter $c_t$ evaluates to $1$, if all configuration options of the corresponding subset $t \subseteq F$ are selected, and $0$ otherwise. The empty set corresponds to the invariable core functionality of the software system. Singleton subsets correspond to individual features, compound subsets to interactions of features.

	For our DBMS example, we present a fully determined performance-influence model in Equation~\ref{eq:pimodel_dbms}. The first two terms correspond to the features \textsc{Encrpytion} and \textsc{Compression}, respectively. The third term respresents the interaction of both features. The configuration parameter $c_{\textsc{Compr}\land\textsc{Encrypt}}$ evaluates to $1$, if both $c_{\textsc{Encrypt}}$ and $c_{\textsc{Compr}}$ evaluate to $1$.
	That is, $c_{\textsc{Compr}\land\textsc{Encrypt}} \equiv c_{\textsc{Encrypt}} \cdot c_{\textsc{Compr}}$. In the last term, we omit the configuration parameter $c_\emptyset$, as this term represents invariable functionality.

@@

	Prediction models of configuration-dependent performance of this form are not the only possible representation~\cite{guoVariabilityawarePerformancePrediction2013,haDeepPerf2019}, but linear models are easy to interpret~\cite{siegmundPerformanceinfluenceModelsHighly2015}.
	We review extensions and alternatives to linear performance-influence models in Section~\ref{sec:rw}.

	\subsection{Evolution of Performance Influences}
	Performance-influence models describe how features contribute to performance, but they do not allow for understanding configuration-dependent performance changes over time.
	We expand our DBMS example with a temporal dimension, considering a development history of hundred commits.
	In particular, we assume the following three events. (1) At commit 25, feature \textsc{Compression} has been modified, increasing the execution time; feature \textsc{Encryption} is introduced at commit 25 and can be combined with \textsc{Compression}. (2) At commit 50, the interaction of \textsc{Compression} and \textsc{Encryption} changes, resulting in an increased execution time. (3) At commit 80, the core functionality of the DBMS is refactored, which decreased execution time for all variants.
	\begin{figure}[ht!]
		%		\begin{subfigure}{\linewidth}
		\centering
		% include first image
		\includegraphics[width=\linewidth]{images/dbms_performance.eps}
		\caption{Performance of 4 variants with 3 change points}
		\label{fig:dbms_performance}
	\end{figure}

	The performance histories of the four valid variants in Figure~\ref{fig:dbms_performance} exhibit three change points: commits 25, 50, and 80. These change points, however, do not affect all configurations.
	For  instance, execution time decreases for all configurations at commit 80, but increases at commit 50 only for one configuration.
	Given this example, the target outcomes of our approach are (1) the locations of the three change points (commits 25, 50, and 80),  (2) the association of such commits to the features and interactions (\textsc{Compr}, $\text{\textsc{Compr}}\ \land\ \text{\textsc{Encrypt}}$), and the invariable functionality, respectively.

@@

	\subsection{Taming Complexity}
	A naive approach for identifying change points in the evolution of configurable software systems is to simply combine existing work on performance modeling of commit histories and performance prediction of software configurations.

	For our example, we could measure all variants for each commit building a performance-influence model per commit.
	Having 4 configurations and 100 commits, this would result in 400 measurements.
	Since the number of configurations grows exponentially with the number of features, we end up with $2^n$ times $T$ measurements where $n$ is the number of features and $T$ is the number of commits.
	Clearly, this does not scale along both dimensions: Already a few features would render even small commit histories intractable, and even small configuration spaces make this approach infeasible given realistic commit histories of a few thousands of commits.

@@

	Based on the introductory DBMS example (cf. Figure~\ref{fig:dbms_performance}), we illustrate our sampling strategy in Figure~\ref{fig:sampling plan}. Here, each box depicts a possible measurement (i.e., a pair of a configuration and a commit). Black bars represent change points hidden in the software system‘s performance histories (i.e., measurements immediately before and after a change point are dissimilar). The current state of the sample set comprises all measurements that are filled in grey.

	To include new measurements into the sample set, we consider the following situations. On the one hand, our sample set might already hint to some possible change points and associated configuration options. On the other hand, our sample set might be too sparse, such that we cannot infer some change points yet. To address this trade-off, we devise two strategies, exploration and exploitation, both of which address both configurations and commits.

	 For \emph{exploitation}, we sample new measurements to verify guesses based on the current state. That is, in Figure~\ref{fig:sampling plan}, we might include the measurements depicted by black-filled symbols~($\starletfill$/$\pentagofill$). Black-filled stars~($\starletfill$) represent measurements that are of interest because we have already identified one change point for configuration \textsc{Compression}. We exploit this knowledge to test further configuration options involved in this change point. The black-filled pentagons~($\pentagofill$) represent measurements that are interesting, because we have identified another early change point for \textsc{Compression}, but within a rather broad range of commits. Thus, we include further measurements from that range to narrow down the possible range for this particular change point.

	 For \emph{exploration}, we might include the measurements denoted by white symbols~($\starlet$/$\pentago$). White stars~($\starlet$) represent measurements that explore new configurations. In our example, the interaction between \textsc{Compression} and \textsc{Encryption} is the only configuration left to be measured. We include several commits of a particular new configuration to unveil possible performance variation that indicates possible change points. White pentagons ($\pentago$) represent measurements that are of interest because they  increase the overall measurement coverage. We include measurements from large intervals of not yet measured commits, since possible change points can be hidden there.

	These four strategies (exploration and exploitation of configurations and commits) prioritize measurements to be included next into the sample set. Adaptive sampling techniques have been successfully applied to obtain both performance-influence models~\cite{siegmundPredictingPerformanceAutomated2012,sarkarCostEfficientSamplingPerformance} and performance histories~\cite{muhlbauer_accurate_2019} before. However, it is unclear whether fewer measurements are sufficient to assess the performance-influence of configuration options and interactions with respect to version changes. This is what we address in this work.

@@

	\section{An algorithm for change point detection}\label{sec:cpmodel}
	We propose an algorithm to detect substantial shifts in the performance of software configurations and associate them to individual commits and options or interactions. We lay out our approach as an iterative search across the commit history and configuration space.

@@

	Subsequently, our algorithm estimates a \emph{candidate solution}, which is a set of pairs of a commit and a configuration option. 
	Each of such pairs describes the estimated involvement of a configuration option in the shift of performance. 
	Interactions are conceived as multiple tuples with identical commits. That is, if a commit occurs in multiple tuples, each with different configuration options, this indicates that the shift arises from an interaction among two or more options. 
	Henceforth, we will refer to such pairs as \emph{associations}.

	After obtaining one candidate solution per search iteration, we augment the sample set of measurements with regard to two objectives: exploration and exploitation. 
	\emph{Exploration} aims at including previously unseen commits and configurations to improve coverage of the search space. 
	\emph{Exploitation} aims at including measurements in the sample set that, based on the previous candidate solution, may increase confidence in associations or rule out false positives (i.e., commits falsely identified as change points or options falsely associated with a commit). 
	As for exploitation, we make an informed decision of which measurements are to be included, whereas exploration is agnostic of previous candidate solutions.

	As each iteration yields a candidate solution, we keep track of associations in a \emph{solution cache} and repeatedly update the confidence of associations. 
	The rationale is not to lose previously identified change points, but at the same time, allow for removing identified changes points that are likely false positives due to new measurements.
	Some associations, especially in the beginning, might be influenced by sampling bias and can be removed if successive iterations do not repeatedly revisit these associations. 
	The algorithm terminates if the solution cache does not change for a number of iterations or a maximum number of iterations/measurements has been reached.
	In what follo, we present the steps of our approach in detail.

	\begin{figure}[ht]
		\includegraphics[width=\linewidth]{images/diagram_april.pdf}
		\caption{Overview of our approach:  \textcircled{1} Selecting training data across configuration space and version history, \textcircled{2} estimating the change point probability distribution per configuration, \textcircled{3} estimating change points as candidate solutions, \textcircled{4} adaptively augmenting the training set, and \textcircled{5} updating the solution cache and evaluating termination.}
		\label{fig:overview}
	\end{figure}

	\subsection{Initialization and Sampling}\label{sec:initialization}
	The first step of our approach is to select a sample set of performance measurements. 
	We select a relatively small, but fixed number of configurations, $n_\text{configurations}$. For each configuration, we select a fixed percentage of commits, $r_\text{commits}$, and assess their respective performance.
	The initial sample set in this setup is kept small and may not represent all relationships between configuration options and performance evolution. 
	The rationale is that, to make our approach more scalable, we explore the combined search space and refine the temporal and  spatial (i.e., configuration-related) resolution where necessary.
